---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading the data
```{r}
file.location<-paste(getwd(),"/activity.csv", sep="")
activity<-read.csv(file.location, 
                   header=TRUE, 
                   colClasses=c("integer", "Date", "integer"))
```

## What is mean total number of steps taken per day?

```{r, message=FALSE}
library(dplyr)
```

```{r}
daily.activities<-select(activity, 2:1)
steps.per.day<-aggregate(daily.activities$steps, 
                         by=list(daily.activities$date), 
                         FUN=sum)
names(steps.per.day)<-c("date", "steps")
mean.per.day<-mean(steps.per.day$steps, na.rm=TRUE)
median.per.day<-median(steps.per.day$steps, na.rm=TRUE)
plot(y=steps.per.day$steps, x=steps.per.day$date, 
     ylab="Steps (total)", xlab="Day", main="Histogram of steps per day",
     lwd=5, col="red", type="h")
```

The daily average is `r mean.per.day`. 
The daily median is `r median.per.day`.  


## What is the average daily activity pattern?

```{r, tidy=TRUE}
interval.activities<-select(activity, -2)
interval.activities<-na.exclude(interval.activities)
steps.per.interval<-aggregate(interval.activities$steps, 
                              by=list(interval.activities$interval), 
                              FUN=mean)
names(steps.per.interval)<-c("interval", "steps")
plot(y=steps.per.interval$steps, x=steps.per.interval$interval,
     ylab="Steps (average)", xlab="Interval (x5min)", 
     main="Average daily pattern", 
     lwd=1, col="blue", type="l")
max.interval<-steps.per.interval$interval[steps.per.interval$steps==
                                              max(steps.per.interval$steps)]
```

The most active interval was `r max.interval`.

## Imputing missing values

```{r}
missing.data<-sum(is.na(activity))
```
The number rows with NAs was `r missing.data`.


Filling in the data by adding previously calculated means for every interval instead of NA.
```{r}
na.activity<-filter(activity, is.na(steps))
compensated.activity<-left_join(select(na.activity, 2:3), 
                                steps.per.interval, 
                                by="interval")
compensated.activity <- select(compensated.activity, c(3,1,2)) 
compensated.activity<-rbind(compensated.activity, 
                            filter(activity, !is.na(steps)))

compensated.daily.activities<-select(compensated.activity, 2:1)
compensated.daily.activities <- arrange(compensated.daily.activities, date)
compensated.steps.per.day<-aggregate(compensated.daily.activities$steps,
                                     by=list(daily.activities$date), 
                                     FUN=sum)
names(compensated.steps.per.day)<-c("date", "steps")
compensated.mean.per.day<-mean(compensated.steps.per.day$steps, na.rm=TRUE)
compensated.median.per.day<-median(compensated.steps.per.day$steps, na.rm=TRUE)
plot(y=compensated.steps.per.day$steps, x=compensated.steps.per.day$date,
     ylab="Steps (total)", xlab="Day", 
     main="Histogram of steps per day (compensated)",
     lwd=5, col="red", type="h")
```

The *compensated* daily average is `r compensated.mean.per.day`. The *compensated* daily median is `r compensated.median.per.day`.

```{r}
difference.mean.per.day<-compensated.mean.per.day-mean.per.day
difference.median.per.day<-compensated.median.per.day-median.per.day
```

The difference in daily averages is `r difference.mean.per.day`.
The difference in daily medians is `r difference.median.per.day`

## Are there differences in activity patterns between weekdays and weekends?
Yes, we can see some differences on the plot - the main difference seems to be that the active period occurs earlier on weekdays.
```{r panel plot, fig.height=7}
week.days<-c("Monday","Tuesday", "Wednesday","Thursday","Friday")
week.ends<-c("Saturday","Sunday")
patterns.activity<-mutate(compensated.activity, days=factor(weekdays(date)))

week.days.activity<-subset(patterns.activity, days %in% week.days)
week.days.steps.per.interval<-aggregate(week.days.activity$steps,
                                        by=list(week.days.activity$interval), 
                                        FUN=mean)
names(week.days.steps.per.interval)<-c("interval", "mean")

week.ends.activity<-subset(patterns.activity, days %in% week.ends)
week.ends.steps.per.interval<-aggregate(week.ends.activity$steps, by=list(week.ends.activity$interval), FUN=mean)
names(week.ends.steps.per.interval)<-c("interval", "mean")

par(mfrow=c(2,1))
plot(y=week.ends.steps.per.interval$mean, x=week.ends.steps.per.interval$interval,
     ylab="Steps (average)", xlab="Interval (x5min)", main="weekend",
     lwd=1, col="blue", type="l")
plot(y=week.days.steps.per.interval$mean, x=week.days.steps.per.interval$interval,
     ylab="Steps (average)", xlab="Interval (x5min)", main="weekday",
     lwd=1, col="green", type="l")
```

